---
title: "B2 assignment"
author: "Luic Damian"
date: "15/03/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r data and packages, include = FALSE}
set.seed(1982)
#### Packages ####
pacman::p_load(here, tidyverse, ggplot2, car, GGally, emmeans, corrplot, vegan, viridis)

#### Read data ####

### Raw data
data = read.csv(here("data/ETX3_B2_multivariate_data2023.csv"))

### Only species
sp = data[1:20]

### Only environmental variables
env = data[21:25]


```

# Data exploration

## Basic statistics and information 

### Dimensions
```{r}
dim(data)   

```
The data set has 30 objects and 25 variables

### Structure

```{r}
str(data) 
```

Species abundance data consists of integers, environmental variables are numeric, i.e. continuous

### Summary

```{r}
summary(data) 

```

The summary shows the minumum, 1st quartile, median, mean, 3rd quartile and max values of each variable



## Distribution and collinearity of predictors 

```{r}
do_env = ggpairs(data, columns = c(21:25), lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.7)))+
  theme_classic()
do_env
    
```
   
All environmental variables are more or less normally distributed and slightly correlated to each other. It should be
checked whether multicollinearity is present or not.

    
## Collinearity

Checking for collinearity with variance inflation parameter (VIF)

```{r}
vif(lm(spp1 ~ env1 + env2 + env3 + env4 + env5, data = data))
```

All VIF for predictors are far below 5, hence there is no problematic collinearity


## Linearity between predictors (environmental variables) and dependent variables (species)

 Because of readablility only 5 species and the 5 environmental parameters are plotted in one graph 
 
```{r data overview pariwise plots} 
do_sp1_sp5_env = ggpairs(data, columns = c(1:5,21:25), lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.7)))+
  theme_classic()
do_sp1_sp5_env

do_sp6_sp10_env = ggpairs(data, columns = c(6:10,21:25), lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.7)))+
  theme_classic()
do_sp6_sp10_env

do_sp11_sp15_env = ggpairs(data, columns = c(11:15,21:25), lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.7)))+
  theme_classic()
do_sp11_sp15_env

do_sp16_sp20_env = ggpairs(data, columns = c(16:20,21:25), lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.7)))+
  theme_classic()
do_sp16_sp20_env

```


There are several significant linear relationships between species abundances and environmental variables. A linerar model to predict abundances from environental parameters may therefore be useful.

Species abundances are often normally distributed but some are skewed, mostly to the right.


## Correlation plot heatmap

```{r}
corrplot(cor(data), method = "color", type = "lower",  tl.col = "black", tl.srt = 45)

```

In this heatmap all correlations between species abundances and environmental parameters can be seen at one glance.
Correlations of varying degree are present between species abundances and environmental variables


## Distribution of species 16 (relevant species for linear models)
```{r}
hist(data$spp16, breaks = 15, xlim = c(0,50), xlab = "Abundance of species 16", main = "Distribution of abundance of species 16 per measuring site")
```

Approximately normally distributed


```{r}
do_sp16_env = ggpairs(data, columns = c(16,21:25), lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.7)))+
  theme_classic()
do_sp16_env
```


There are correlations between abundance of species 16 and the environmental parameters, the correlation between 
spp16 and env4 being significant



# Linear models using stepwise model selection

## Intercept and full model

To apply stepwise model selection, the intercept model (model without any environmental predictor) and the
full model (model with all environmental parameters as predictors) must be specified

```{r}
intercept_model = lm(spp16 ~ 1, data = data)


full_model = lm(spp16 ~ env1 + env2 + env3 + env4 + env5, data = data)

```


## Not including interaction terms
### Forward stepwise model selection (no interaction) 

```{r}
for_lm = step(intercept_model, scope=formula(full_model), directions = "forward")
for_lm$coefficients

```

Resulting model: $$spp16 = -0.327*env4 + 77$$

### Backward stepwise model selection (no interaction) 

```{r}
back_lm = step(full_model, scope=formula(full_model), directions = "backward")
back_lm$coefficients
```


Resulting model: $$spp16 = -0.327*env4 + 77$$


### Both ways stepwise model selection (no interaction) 

```{r}
both_lm = step(intercept_model, scope=formula(full_model), directions = "both")
both_lm$coefficients
```


Resulting model: $$spp16 = -0.327*env4 + 77$$


*All stepwise model selection methods without interactions resulted in the same model*
*using only the predictor env4 to predict abundance of species 16*

### Model diagnosis
```{r}
plot(for_lm)
```


Assumptions are fulfilled:   
- residuals are homogeneous (Homoscedasticity)
- Normal distribution of error
- There are no infuential outliers
- Observations should be independent, but  data source is unknown
- Linearity has been checked in line 96
- Multicollinearity has been excluded in line 54

```{r}
summary(for_lm)

```


The relationship between env4 and spp16 is highly significant (p < 8*10^-5)
But only about 43 % of variance are explained by the model.


## Models with interaction terms

This time the full model includes interactions

```{r}
intercept_model = lm(spp16 ~ 1, data = data)


full_model_int = lm(spp16 ~ env1 * env2 * env3 * env4 * env5, data = data)
summary(full_model_int)
```



With all interactions included in the full model, there are more predictors than observations.
This is impossible to calculate.
To demonstrate the inclusion of interaction terms nevertheless, the parameter env5 is simply removed
from the model.

```{r}
full_model_int = lm(spp16 ~ env1 * env2 * env3 * env4, data = data)
summary(full_model_int)

```

Now stepwise model selection can be done:

### Forward stepwise model selection (with interaction) 
```{r}
for_lm_int = step(intercept_model, scope=formula(full_model_int), directions = "forward")
for_lm_int$coefficients
```



Resulting model: $$spp16 = -0.327*env4 + 77.552$$ 
(only env4 as predictor)


### Backward stepwise model selection (with interaction) 

```{r}
back_lm_int = step(full_model_int, scope=formula(full_model_int), directions = "backward")
back_lm_int$coefficients
```


Resulting model contains the predictors *env1, env2, env3, env4, env2:env3, env1:env4, env2:env4,* 
*env3:env4 and env2:env3:env4*

### Both ways stepwise model selection (with interaction) 

```{r}
both_lm_int = step(intercept_model, scope=formula(full_model_int), directions = "both")
both_lm_int$coefficients

```


Resulting model: $$spp16 = -0.327*env4 + 77.552$$
(only env4 as predictor), same model as forward stepwise selection


Forward and both directions stepwise selection resulted in a sparce model containing only env4 as a predictor
Backward stepwise selection included env1, env2, env3, env4, env2:env3, env1:env4, env2:env4, 
env3:env4 and env2:env3:env4

### Diagnosis and interpretation ####

#### Forward and both directions stepwise model selection

Both resulted in the same model
```{r}
plot(for_lm_int)

```

Assumptions for forward and both direction stepwise selection models are fulfilled:   
- residuals are homogeneous (Homoscedasticity)
- Normal distribution of error
- There are no infuential outliers
- Observations should be independent, but  data source is unknown
- Linearity has been checked in line 96
- Multicollinearity has been excluded in line 54

#### Backward model selection

```{r}
plot(back_lm_int)

```

Assumptions for backward stepwise selection model are not completely fulfilled: 
- residuals are only more or less homogeneous (Homoscedasticity)
- Error is normally distributed although not as clearly as the previous models
- There are at least two infuential outliers
- Observations should be independent, but data source is unknown
- Linearity has been checked in data exploration
- Multicollinearity is present due to interaction terms (see below) 

Checking for multilinearity in the linear model with interaction terms:

```{r}
vif(lm(spp1 ~ env1 * env2 * env3 * env4, data = data))
```
Very high VIFs -> multicollinearity

```{r}
vif(lm(spp1 ~ env1 + env2 + env3 + env4, data = data)) # Low VIFs in the model without interactions

```
Low VIFs in the model without interactions

```{r}
vif(lm(spp1 ~ env1:env2 + env2:env3 + env1:env4 + env2:env4 + env3:env4 + env1:env2:env3 +env1:env2:env4 + env1:env3:env4 + env2:env3:env4 + env1:env2:env3:env4, data = data))
```
High VIFs in model with only interactions


#### Interpretation forward and both directions model

```{r}
summary(for_lm_int)

```

The relationship between env4 and spp16 is highly significant (p < 8*10^-5)
But only about 43 % of variance are explained by the model (or 41 % when considering adjusted R^2)


#### Interpretation backward stepwise selection model
```{r}
summary(back_lm_int)

```


There is a significant relationship between model predictors and abundance of species 16 (p = 0.022 < 0.05)
R^2 is higher than the sparse model including only env4, however adjusted R^2 is lower due to the number 
of predictors. The relationship is also less significant.
Overall the sparse model including only env4 (and no interactions) would be preferred as it explains a
considerable degree of variance using only one predictor and very low p-value



# Nonmetric multidimensional scaling (NMDS) 

Calculating NMDS using "Bray-Curtis-dissimilarity" as a distance metric

```{r, echo = FALSE}
nmds = metaMDS(comm = data[,1:20],
               distance = "bray",
               try = 100)

```


Plotting the NMDS graph

```{r}
plot(nmds)
orditorp(nmds,display="species",col="red",air=0.01, cex = 0.8) # indicating species points in graph 
orditorp(nmds,display="sites",cex=1.25,air=0.01) # indicating site points in graph
```

A multitude of information may be gathered from the NMDS graph. Some of the insights might be:
1.) site 27 and 2 are very different
2.) Sites 13, 5, 14, 11, 9 and 17 cluster together and seem to be rather similar
3.) Species 8 and species 3 seem to be the most similar species
4.) Species 2 and species 13 are the most different


